# Caddo bottle asymmetry

In studies of ceramic morphology, an analysis of bilateral/object symmetry can yield useful insights related to _design intent_, provided it is assumed that the goal of the potter was to achieve a symmetrical vessel. Within that framework, deviation from bilateral symmetry (fluctuating asymmetry) can serve as a useful measure of _stability_, _skill_ (do Caddo bottles become more symmetrical through time?) and/or _differential production_ (are the bottles from one community of practice more symmetrical than another?), and both measures have utility in spatial and temporal comparisons.

## Load packages + data

```{r load.data.asymm, echo = TRUE}
# load packages

# devtools::install_github("geomorphR/geomorph", ref = "Stable", build_vignettes = TRUE)
library(here)
library(geomorph)
library(tidyverse)
library(wesanderson)

# read GM data
source('readmulti.csv.R')

# read .csv files
setwd("./data")
filelist <- list.files(pattern = ".csv")
coords <- readmulti.csv(filelist)
setwd("../")

# read qualitative data
qdata <- read.csv("qdata.csv", header = TRUE, row.names = 1)
qdata <- qdata[match(dimnames(coords)[[3]],rownames(qdata)),]
land.pairs<-read.csv("asymmetry1.csv",header=TRUE)
ind<-rep(1:length(filelist))

# print qdata
knitr::kable(qdata, align = "lcccc", caption = "Attributes included in qdata.")
```

## Generalised Procrustes Analysis

```{r assym.gpa, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# gpa for bilateral (object) asymmetry
res.bilat <- bilat.symmetry(coords,
                          ind = ind, 
                          object.sym = TRUE,
                          land.pairs = land.pairs,
                          RRPP = TRUE,
                          iter = 9999,
                          print.progress = FALSE)
summary(res.bilat)
```

```{r symm.shape, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Symmetric shape component plotted in 2D."}
# symmetric shape
symm.shape <- res.bilat$symm.shape[,1:2,]

# plot all specimens in 2D
plotAllSpecimens(symm.shape)
```

```{r asymm.shape, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE, fig.cap="Asymmetric shape component plotted in 2D."}
# asymmetric shape
asymm.shape <- res.bilat$FA.component[,1:2,]

# plot all specimens in 2D
plotAllSpecimens(asymm.shape)
```

## Principal Components Analysis

```{r pca.asymm, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# geomorph data frame
gdf <- geomorph.data.frame(shape = res.bilat$symm.shape, 
                           ashape = res.bilat$FA.component,
                           time = qdata$time,
                           geo = qdata$geo)

# principal components analysis (symmetric)
pca1 <- gm.prcomp(res.bilat$symm.shape)
summary(pca1)

# set plot parameters
pch.gps.time <- c(15,17)[as.factor(qdata$time)]
col.gps.time <- wes_palette("Moonrise2")[as.factor(qdata$time)]
col.hull <- c("#C27D38","#798E87")

# plot pca1 (symmetric)
pc.plot1 <- plot(pca1, 
                 asp = 1,
                 pch = pch.gps.time,
                 col = col.gps.time)
                    shapeHulls(pc.plot1, 
                             groups = qdata$time,
                             group.cols = col.hull)
                    
# principal components analysis (asymmetric)
pca2 <- gm.prcomp(res.bilat$asymm.shape)
summary(pca2)

# plot pca2 (asymmetric)
pc.plot2 <- plot(pca2, 
                 asp = 1,
                 pch = pch.gps.time,
                 col = col.gps.time)
                    shapeHulls(pc.plot2, 
                             groups = qdata$time,
                             group.cols = col.hull)
```

## Define models

A residual randomization permutation procedure (RRPP; n = 10,000 permutations) was used for all Procrustes ANOVAs [@RN1655;@RN11775], which has higher statistical power and a greater ability to identify patterns in the data should they be present [@RN1719]. To assess whether shape changes with size (allometry), and differs by group (region), Procrustes ANOVAs [@RN1749] were also run that enlist effect-sizes (zscores) computed as standard deviates of the generated sampling distributions [@RN1756]. 

_Hypothesis 1: There is a difference in the symmetric shape of Caddo bottles through time._

```{r def.mod.1.asymm, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# symmetric shape as a function of time
fit.time <- procD.lm(shape ~ time,
                     data = gdf,
                     print.progress = FALSE,
                     iter = 9999)

# asymmetric shape as a function of time
fit.time.asymm <- procD.lm(ashape ~ time,
                           data = gdf,
                           print.progress = FALSE,
                           iter = 9999)
```

_Hypothesis 2: There is a difference in the symmetric shape of Caddo bottles by community of practice._

```{r def.mod.2.asymm, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# symmetric shape as a function of geography
fit.geo <- procD.lm(shape ~ geo,
                     data = gdf,
                     print.progress = FALSE,
                     iter = 9999)

# asymmetric shape as a function of geography
fit.geo.asymm <- procD.lm(ashape ~ geo,
                           data = gdf,
                           print.progress = FALSE,
                           iter = 9999)
```

## Procrustes ANOVA (H1)

```{r h1 asymm, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# ANOVA: do symmetric Caddo bottle shapes differ by time?
## potential differences in design intent?
anova(fit.time)

# ANOVA: do asymmetric Caddo bottle shapes differ by time?
## potential differences in design intent?
anova(fit.time.asymm)
```

## Procrustes ANOVA (H2)

```{r h2 asymm, out.width = "100%", dpi = 300, echo=TRUE, warning=FALSE}
# ANOVA: do symmetric Caddo bottle shapes differ by geography?
## potential differences in skill?
anova(fit.geo)

# ANOVA: do asymmetric Caddo bottle shapes differ by geography?
## potential difference in skill?
anova(fit.geo.asymm)

# end of code
```
